---
title: '2: The tidyverse'
author: "Justin Sulik"
output: ioslides_presentation
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(knitr)
```

## Modern R: the Tidyverse

- The tidyverse is just a collection of packages
   - tidyr
   - dplyr
   - ggplot
   - ...

- You can think of it as a manual car vs an automatic car

## Modern R: the Tidyverse

Some reasons we're using it:

- streamlining your workflow
- making the code easier to read
- great for manipulating data
 
## Basic concept: the pipe

Often we need to do a series of operations on something

- either create a bunch of new values, step by step:

```{r, out.width='100%', eval=F}
input
output1 <- do_something(input)
output2 <- do_something_else(output1)
...
```

- or embed the functions

```{r, out.width='100%', eval=F}
output2 <- do_something_else(do_something(input))
```

## Example 1: no pipes

<div style="float: left; width:50%">
```{r, out.width='100%'}
set.seed(123)
input <- rnorm(100, 5, 3)
step1 <- mean(input)
step2 <- round(step1, 2)
step2
```
</div>

<div style="float: right; width:50%">
```{r, out.width='100%'}
set.seed(123)
round(mean(rnorm(100,5,3)),2)
```
</div>

## Example 2: pipes

```{r, eval=F}
input >
  do_something >
  do_something_else
```

Except that custom operators in R usually have %...% around them

```{r, eval=F}
input %>%
  do_something %>%
  do_something_else
```

## Example 2: pipes

```{r}
set.seed(123)

rnorm(100, 5, 3) %>%
  mean %>%
  round(2)
```

## Let's get some data

```{r}
sci_denial <- read_csv("data/science_denial.csv")
demographics <- read_csv("data/demographics.csv")
```

## Check out the data

```{r}
head(demographics)
```

## Check out the data

```{r}
summary(demographics)
```

## Check out the data

```{r}
str(demographics)
```

## Check out the data

```{r}
names(demographics)
```


## Tidyverse verbs: Filtering and selecting

```{r, eval=F}
filter(data, ...conditions for rows...)

select(data, ...column identifiers...)
```

## Select columns {.smaller}

```{r}
demographics %>% 
  select(participant, age)
```

## Select columns {.smaller}

```{r}
demographics %>% 
  select(-participant, -age)
```

## Select columns {.smaller}

How to get a range of columns?

```{r}
sci_denial %>% 
  select(credibility_0:credibility_5)
```

## Select columns {.smaller}

How to get a range of columns?

```{r}
sci_denial %>% 
  select(starts_with("credibility"))
```

## Filter rows {.smaller}

```{r}
demographics %>% 
  filter(english==1)
```

## Filter rows {.smaller}

 & = and; | = or; ! = not

```{r}
demographics %>% 
  filter(english==1 | education==6)
```

## Filter rows {.smaller}

```{r}
demographics %>% 
  filter(english==1, education==6)
```

## Filter rows {.smaller}

```{r}
demographics %>% 
  filter(english==1, education==6, age < 60)
```

## Gather (wide to long)

```{r}
sci_denial %>% 
  gather(item, response, credibility_0:sciKnowledge_9)
```

## Gather (wide to long)

```{r}
sci_denial %>% 
  gather(item, response, -participant)
```

## Gather (wide to long)

- Opposite is "spread"

- Why long data?
   - easier to summarise (in 5 minutes)
   - easier to run fancy regressions (tomorrow)
 
## Separate (split column data)

- Opposite is "unite"

```{r}
sci_denial %>% 
  gather(item, response, -participant) %>% 
  separate(item, c("scale", "question"))
```

## Group_by and summarise

```{r}
demographics %>% 
  group_by(gender) %>% 
  summarise(age=mean(age))
```

## Group_by and summarise

```{r}
demographics %>% 
  group_by(gender) %>% 
  summarise(age=mean(age),
            n=n())
```

## Group_by and summarise

```{r}
demographics %>% 
  group_by(gender) %>% 
  summarise_if(is.numeric, mean)
```

## Exercise

Use `group_by` and `summarise` on the `sci_denial` data, to calculate an average score for each participant for scale.

(if you finish soon: use `?spread` to try return this to wide format)

## Mutate

Change a column or create a new column

```{r}
demographics %>% 
  mutate(decade = round(age/10, 0))
```

! Be *very* cautious about loosing information in this way

## Mutate

Change a column or create a new column

```{r}
demographics %>% 
  mutate(senior = ifelse(age>=60, TRUE, FALSE))
```

## Mutate

Change a column or create a new column

```{r}
demographics %>% 
  mutate(gender = ifelse(!gender %in% c("male", "female", "na"),
                         gender,
                         "nb")) 
```

**ALWAYS double check it has worked as expected. Think of a way to check the above mutation**

## Exercise

Open the `personality` data file

Calculate a score for each personality construct (A = agreeableness; C = Conscientiousness; E = Extraversion; N = Neuroticism; O = Openness). 

Put that score in separate columns, one for each construct. 

## Joining dataframes by common columns {.smaller}

```{r}
demographics %>% 
  left_join(sci_denial) %>% 
  head(5)
```

**Think of two ways to join tables that -- for whatever reason --  have 'participant' in one table and 'ID' in the other table, both referring to the same thing**

## Joining dataframes by common columns {.smaller}

```{r, message=F, warning=F}
demographics %>% 
  left_join(
    sci_denial %>% 
      gather(item, response, -participant) %>% 
      separate(item, c("scale", "question")) %>% 
      group_by(participant, scale) %>% 
      summarise(score = mean(response)) %>% 
      spread(scale, score)
  ) %>% 
  head(5)
```

## Other useful functions

```{r}
beads <- read_csv("data/beads.csv")
```

<video width="320" height="240" controls>
  <source src="images/demo.mp4" type="video/mp4">
</video>

## Other useful functions

lag - return value from previous row

```{r}
beads %>% 
  mutate(previous_bead = lag(bead),
         same = ifelse(bead==previous_bead, T, F))
```

**What verb that we've learned can help ensure each participant's first "same" is NA?**

## Other useful functions {.smaller}

cumulative functions - e.g. cumsum

```{r}
beads %>% 
  mutate(bead = ifelse(bead=="white", 1, 0)) %>% 
  group_by(participant) %>% 
  mutate(white_beads = cumsum(bead))
```

## Other useful functions

arrange - ordering columns

```{r}
demographics %>% 
  arrange(age) %>% 
  head(5)
```

**What if we want oldest to youngest?**

## Summary

- Don't be afraid to mess around with your data

- Just check at every stage that it's doing what you expect

- This is tough to remember: 
    - copy+paste+adapt
    - needs practice

## Exercise (if time remaining)

What if the last 4 items of the scienceKnowledge scale measure something specific: science denial?

Update previous code to generate two separate scores for science knowledge and science denial. 

