---
title: '5: Increasing complexity: interactions and polynomials'
author: "Justin Sulik"
output: 
    ioslides_presentation:
        smaller: true
---

```{r, include=F}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(knitr)
library(faux)
library(ggpubr)

theme_set(theme_bw())
```

## So far...

- We have been considering main/linear effects
    - What's the effect of x1 on y? Of x2?
    - Yes, they can change due to collinearity
    - But it's still a main effect: 
    - For a given dataset, we modelled the effect of x as unchanging
    
## Up next...
    
- But how might the effect of x change?
    - The effect of x gets higher/lower for higher/lower values of x
    - The effect of x1 gets higher/lower for higher/lower values of x2
    
## Up next...
    
- But how might the effect of x change?
    - The effect of x gets higher/lower for higher/lower values of x $\longrightarrow$ **higher-order polynomial**
    - The effect of x1 gets higher/lower for higher/lower values of x2 $\longrightarrow$ **interaction**

## Quadratic (increasing)

```{r}
set.seed(123)
data_quad <- data.frame(x=rnorm(100,5,2),
                        beta1=rnorm(100,2,2),
                        beta2=rnorm(100,1,1)) %>% 
  mutate(y=beta1*x + beta2*x^2)

ggplot(data_quad, 
       aes(x=x, y=y)) + 
  geom_point(alpha=0.4) + 
  stat_smooth(method=lm, 
              formula = y ~ x + I(x^2))
```

## Quadratic (decreasing)

```{r}
set.seed(1234)
data_quad <- data.frame(x=rnorm(100,2,2),
                        beta1=rnorm(100,2.3,0.5),
                        beta2=rnorm(100,-0.1,0.01)) %>% 
  mutate(y=beta1*x + beta2*x^2)

ggplot(data_quad, 
       aes(x=x, y=y)) + 
  geom_point(alpha=0.4) + 
  stat_smooth(method=lm, 
              formula = y ~ x + I(x^2)) 
```

## Quadratic

$y \sim b_0 + b_1*x + b_2*x^2$

Can you think of any reasonable examples?

What's the harm in ignoring it?

When should I care? THEORY (+ fit)

## Model comparison {.smaller}

Try to improve fit, but only if it's worth it

BIC = one measure of "worth it"; prefers more parsimonious models

```{r, echo=T}
mod_linear <- lm(y~x,data_quad)
summary(mod_linear)
```

## Model comparison {.smaller}

Try to improve fit, but only if it's worth it

BIC = one measure of "worth it"; prefers more parsimonious models

```{r, echo=T}
mod_quadratic <- lm(y~x+I(x^2),data_quad)
summary(mod_quadratic)
```

## Model comparison

```{r}
BIC(mod_linear, mod_quadratic)
```

## Model comparison {.smaller}

Lower BIC = better

Difference in BIC  Claim
-----------------  ---------
<2.0               Weak
2.0–5.9            Positive 
6.0–9.9            Strong 
≥10.0              Very strong
-----------------  ---------

[https://sites.stat.washington.edu/raftery/Research/PDF/socmeth1995.pdf](https://sites.stat.washington.edu/raftery/Research/PDF/socmeth1995.pdf)

## Interactions

- What is an interaction?
  - Verbal
  - Visual
  - Math
- Re-leveling factors/centering variables
- Power
- When you *should* include an interaction

## Example without an interaction

- Imagine you are developing a new feed for salmon farming
- Let's say on average it produces a 300g increase in fish size

```{r fish_data1, cache=T}
fish_data <- data.frame(gender=rep(c("male", "female"), 1000),
         feed = rep(c("control", "new"), each=1000)) %>% 
  mutate(weight = rnorm(2000, 2000, 400),
         weight = ifelse(gender=="male", weight+400, weight),
         weight = ifelse(feed=="new", weight+300, weight))
```

```{r, out.width="70%"}
fish_data %>% 
  ggplot(aes(x=feed, y=weight)) + 
  geom_point(position="jitter",
             alpha=0.4) +
  stat_summary(fun=mean,
               geom="point",
               color="red",
               shape=18,
               size=10) + 
  labs(y="weight (g)",
       title="Effect of feed type")
```

## Example without an interaction

- But then you remember that male fish are 400g bigger on average
- Want to check you've controlled for that

```{r, out.width="70%"}
fish_data %>% 
  ggplot(aes(x=gender, y=weight)) + 
  geom_point(position="jitter",
             alpha=0.4) +
  stat_summary(fun=mean,
               geom="point",
               color="red",
               shape=18,
               size=10) + 
  labs(y="weight (g)",
       title="Effect of gender")
```

## Example without an interaction

- So we have 2 main effects:
    - new feed: +300g
    - gender male: +400g
    
- Assuming these *effects* are independent of each other, and given that an average female fish on the regular feed weighs 2000g, what would you expect the average weight to be for:
    - a female on the new feed?
    - a male on the regular feed?
    - a male on the new feed?
  
**Note: this is not about a correlation betw. gender and feed. What would that even mean?**

## Example without an interaction

- weight = $b_0 + b_1*gender + b_2*feed$
    - (imagine female = 0, male = 1; control=0, new=1)
- weight = $2000 + 400*gender + 300*feed$

```{r, echo=T}
mod1 <- lm(weight ~ gender + feed,
           data=fish_data)
coef(mod1)
```


## Example without an interaction

- What's the name of this shape?

```{r}
fish_data %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender,
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.4) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) 
```

## Example without an interaction

- What's the name of this shape?

```{r}
fish_data %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender,
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.1) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) +
  geom_segment(x = "control", y=2000,
               xend = "control", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2700,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2700,
               xend = "new", yend = 2300,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2300,
               xend = "control", yend = 2000,
               color = "black",
               size = 2)
```


## What is an interaction?

- Imagine the new feed had a different effect on males and female fish
- E.g.,  +400g for females; +200g for males
- Thus, the effect of one variable depends on the value of another 

## What is an interaction?  


```{r fish_data2, cache=T}
fish_data_interaction <- data.frame(gender=rep(c("male", "female"), 1000),
         feed = rep(c("control", "new"), each=1000)) %>% 
  mutate(weight = rnorm(2000, 2000, 200),
         weight = ifelse(gender=="male", weight+400, weight),
         weight = ifelse(feed=="new", weight+400, weight), 
         weight = ifelse(feed=='new' & gender=='male', weight-200, weight))
```

```{r}
fish_data_interaction %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender, 
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.1) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) +
  geom_segment(x = "control", y=2000,
               xend = "control", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2600,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2600,
               xend = "new", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2400,
               xend = "control", yend = 2000,
               color = "black",
               size = 2)
```

## What is an interaction?  

```{r}
fish_data_interaction %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender, 
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.1) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) +
  geom_segment(x = "control", y=2000,
               xend = "control", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2600,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2600,
               xend = "new", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2400,
               xend = "control", yend = 2000,
               color = "black",
               size = 2) +
  annotate(geom="text",
           label="+400",
           x=1.5,
           y=2100) + 
  annotate(geom="text",
           label="+200",
           x=1.5,
           y=2600)
```

## What is an interaction?  


```{r}
fish_data_interaction %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender, 
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.1) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) +
  geom_segment(x = "control", y=2000,
               xend = "control", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2600,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2600,
               xend = "new", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2400,
               xend = "control", yend = 2000,
               color = "black",
               size = 2) +
  annotate(geom="text",
           label="+400",
           x=0.9,
           y=2200) + 
  annotate(geom="text",
           label="+200",
           x=2.1,
           y=2500)
```

## What is an interaction?  | Your turn

- weight = $b_0 + b_1*gender + b_2*feed + b_3*gender*feed$
  - (gender: female = 0, male = 1; feed: control = 0, new = 1)
- Try map the parameters onto by-group means!

## What is an interaction?  | Adding in an interaction

```{r}
fish_data_interaction %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender, 
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.1) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) +
  geom_segment(x = "control", y=2000,
               xend = "control", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2600,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2600,
               xend = "new", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2400,
               xend = "control", yend = 2000,
               color = "black",
               size = 2) + 
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2800,
               color = "grey",
               size = 1) +
  geom_segment(x = "new", y=2800,
               xend = "new", yend = 2600,
               color = "grey",
               size = 1) +
  annotate(geom="text",
           x=0.9,
           y=2200,
           label="+400") + 
  annotate(geom="text",
           x=1.5,
           y=2100,
           label="+400")
```

## What is an interaction? 

```{r}
mod2 <- lm(weight ~ gender*feed, 
           fish_data_interaction)
coef(mod2)
```

```{r}
fish_data_interaction %>% 
  group_by(gender,feed) %>% 
  summarise(m=mean(weight))
```


## What is an interaction?

- Verbally: Are the effects independent? Or does the strength of one effect depend on the value of the other?
- Visually: parallelogram or not?
    - *Always* plot your data!
- Mathematically: $b_1x_1 + b_2x_2 + b_3x_1x_2$

## What is an interaction?

- We had **a reason** to look for an interaction term (mileage can vary)
    - Knowing gender has an effect on size
    - Wanting to know if the feed works on all salmon
    - Wanting to know whether the mechanism by which the feed increases size has anything to do with gender-variable hormones, 
    - etc. 
    
- Your analyses **should** be theory driven

## Centering/relevelling 

What if we exclude the interaction term?

Run a linear regression with just the two main effects (data: `fish_data_interaction`)

## Centering/relevelling | Why did the numbers change?

```{r}
mod3 <- lm(weight~gender+feed, 
           fish_data_interaction)
summary(mod3)
```

## Centering/relevelling | Why did the numbers change?

```{r}
fish_data_interaction %>% 
  ggplot(aes(x=feed, 
             y=weight,
             color=gender, 
             group=gender)) + 
  geom_point(position="jitter",
             alpha=0.1) + 
  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 10) +
  geom_segment(x = "control", y=2000,
               xend = "control", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "control", y=2400,
               xend = "new", yend = 2600,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2600,
               xend = "new", yend = 2400,
               color = "black",
               size = 2) +
  geom_segment(x = "new", y=2400,
               xend = "control", yend = 2000,
               color = "black",
               size = 2) +
  annotate(geom="text",
           label="+400",
           x=0.9,
           y=2200) + 
  annotate(geom="text",
           label="+200",
           x=2.1,
           y=2500) + 
  annotate(geom="text",
           label="+400",
           x=1.5,
           y=2100) + 
  annotate(geom="text",
           label="+200",
           x=1.5,
           y=2600)
```

## Centering/relevelling | Where did my effect go?

New example: you're testing whether a new "cognitive enhancement" drug speeds up reaction times in a decision making task. Control group gets a placebo; experimental group gets the drug. Pre- and post-treatment measures. *Assuming* that the drug works:

- What do you expect to happen to the RT in the control group?
- What do you expect to happen to the RT in the experimental group?

```{r drug_data, cache=T}
N <- 1000
drug_data <- data.frame(id = 1:N,
                        group = rep(c("control", "expt"), each=N/2),
                        baseline = rnorm(N, 1300, 400),
                        change = rnorm(N, -400, 400),
                        nochange = rnorm(N, 0, 400)) %>% 
  mutate(posttreatment = ifelse(group=="control", baseline+nochange, baseline+change)) %>% 
  gather(phase, rt, baseline, posttreatment) %>% 
  select(-change, -nochange)
```

## Centering/relevelling | Where did my effect go?

```{r}
drug_data %>% 
  ggplot(aes(x=phase, y=rt, color=group, group = group)) + 
  geom_point(alpha=0.4, position="jitter") + 
  stat_summary(fun = mean,
               geom = "point",
               size = 10,
               shape = 18) +
  stat_summary(fun = mean,
               geom = "line", 
               size = 2)
```

## Centering/relevelling | Where did my effect go?

Does it look like there's an effect of the drug? Yes!

So do you expect:

- a significant main effect of drug?
- a significant interaction term?

## Centering/relevelling | Where did my effect go?

- Model just the main effects (data = `drug_data`)

## Centering/relevelling | Where did my effect go?

- Model just the main effects (data = `drug_data`)

```{r}
mod4 <- lm(rt ~ group + phase, 
   drug_data)

summary(mod4)
```

## Centering/relevelling | Where did my effect go?

- Add an interaction term

## Centering/relevelling | Where did my effect go?

- Add an interaction term

```{r}
mod5 <- lm(rt ~ group*phase, 
   drug_data)

summary(mod5)
```

## Centering/relevelling | Where did my effect go?

```{r}
drug_data %>% 
  ggplot(aes(x=phase, y=rt, color=group, group = group)) + 
  geom_point(alpha=0.1, position="jitter") + 
  stat_summary(fun = mean,
               geom = "point",
               size = 10,
               shape = 18) +
  stat_summary(fun = mean,
               geom = "line", 
               size = 2) + 
  geom_segment(x = 1.5, 
               y = 1300,
               xend = 1.5,
               yend = 1100,
               color = "black") + 
  geom_segment(x = 1, 
               y = 1300,
               xend = 2,
               yend = 1100,
               color = "black") + 
  labs(title="Model with main effects only")
```

## Centering/relevelling | Where did my effect go?

```{r}
group_means <- drug_data %>% group_by(phase,group) %>% summarise(m=mean(rt)) %>% pull(m)

drug_data %>% 
  ggplot(aes(x=phase, y=rt, color=group, group = group)) + 
  geom_point(alpha=0.1, position="jitter") + 
  stat_summary(fun = mean,
               geom = "point",
               size = 10,
               shape = 18) +
  stat_summary(fun = mean,
               geom = "line", 
               size = 2) + 
  geom_segment(x = 2, 
               y = group_means[3],
               xend = 2,
               yend = group_means[2],
               color = "grey") + 
  geom_segment(x = 1, 
               y = group_means[2],
               xend = 2,
               yend = group_means[2],
               color = "grey") + 
  geom_segment(x = 1, 
               y = group_means[1],
               xend = 1,
               yend = group_means[2],
               color = "black") + 
  geom_segment(x = 1, 
               y = group_means[1],
               xend = 2,
               yend = group_means[3],
               color = "black") + 
  geom_segment(x = 2.01, 
               y = group_means[2],
               xend = 2.01,
               yend = group_means[4],
               color = "black") +
  labs(title="Model with interaction term")
```

## Centering/relevelling | Where did my effect go?

By default:

- No interaction term: population average effect
- Interaction term (effect of $x_1$ depends on $x_2$):
   - $b_1x_1$: when $x_2=0$ 
   - $b_2x_2$: when $x_1=0$ 
   - $b_3x_1x_2$: here's where $x_1$ and $x_2$ can vary
   
## Centering/relevelling | Where did my effect go?

So what does 0 mean here?

By default:

- numeric variable: literal 0
- categorical variable: first alphabetically (female - male; control - new; baseline - posttreatment; control - expt)

So what can you do?

- center your numeric variables
- center/relevel your categorical variables

## Centering | Numeric variables

```{r, echo=T}
drug_data %>% 
  mutate(rt_scaled = scale(rt)) %>% 
  head
```

## Centering | Categorical variables

```{r}
# Center variables (from https://hlplab.wordpress.com/2009/04/27/centering-several-variables/)
myCenter <- function(x) {
  if (is.numeric(x)) { return(x - mean(x, na.rm=T)) }
  if (is.factor(x)) {
    x = as.numeric(x)
    return(x - mean(x, na.rm=T))
  }
  if (is.character(x)) {
    x = as.numeric(as.factor(x))
    return(x - mean(x, na.rm=T))
  }
  if (is.data.frame(x) || is.matrix(x)) {
    m= matrix(nrow=nrow(x), ncol=ncol(x))
    colnames(m)= paste("c", colnames(x), sep="")
    for (i in 1:ncol(x)) {
      m[,i]= myCenter(x[,i])
    }
    return(as.data.frame(m))
  }
}
```


```{r, echo=T}
fish_data_interaction %>% 
  mutate(gender_centered = myCenter(gender)) %>% 
  head
```

## Centering | Categorical variables

```{r}
drug_data %>% 
  mutate(phase_centered = myCenter(phase),
         group_centered = myCenter(group)) %>% 
  lm(rt ~ phase_centered*group_centered,.) %>% 
  summary
```

## Centering | Do I *need* to do this?

No, it depends on your research question

- Fish example: useful to know what the overall effect of gender is
- Drug example: useful to know that there's no difference at baseline, no effect of phase on control, and only a change in for experimental group post treatment

## Releveling | Changing which is the base level (=0)

Imagine we'd called it "before" and "after" instead of "baseline" and "posttreatment"

```{r}
drug_data %>% 
  mutate(phase = ifelse(phase=="baseline", "before", "after")) %>% 
  ggplot(aes(x=phase, y=rt, color=group)) + 
  geom_point(alpha=0.1,
             position = "jitter") + 
  stat_summary(fun = mean,
               geom = "point",
               size = 10,
               shape = 18)
```

## Releveling | Changing which is the base level (=0)

Imagine we'd called it "before" and "after" instead of "baseline" and "posttreatment"

```{r}
drug_data %>% 
  mutate(phase = ifelse(phase=="baseline", "before", "after")) %>% 
  lm(rt ~ phase*group,.) %>% 
  summary
```

## Releveling | Changing which is the base level (=0)

```{r, echo=T}
fish_data_interaction %>% 
  mutate(gender = factor(gender, levels=c("male", "female"))) %>% 
  lm(weight ~ gender,.) %>% 
  summary
```

## Centering/relevelling | Summary

- Which you do (if any) depends on what you're interested in
- This will change the patterns of significance BUT THAT'S FINE

```{r}
drug_data %>% 
  ggplot(aes(x=phase, 
             y=rt,
             group=group)) + 
  geom_point(aes(color=group),
             position="jitter",
             alpha=0.1) + 
  stat_summary(aes(color=group),
               fun=mean,
               geom="point",
               size=10,
               shape=18,
               alpha=0.6) + 
  annotate("segment",
           x = 1.5, y=900,
           xend = 1.5, yend = 1300,
           color = "black",
           size = 1,
           arrow = arrow()) +
  annotate("text",
           x = 1.5, y=1380,
           label = "Center") +
  geom_bracket(xmin = "baseline", xmax = "posttreatment",
               y.position = 2000,
               label = "Relevel") 
```

## Power

What should my sample size be if I need to test for an interaction?

- Big

- VERY INFORMAL rule of thumb: if you need $N$ to detect main effect $b$, might need $4N$ to detect interaction size $b$

- We'll worry more about power tomorrow

## When should you include an interaction?

- If, given your best understanding of the relevant literature, it's plausible that the effect of one variable depends on the value of another
    - e.g., varies by group
- If you are wanting to say an effect in one group is larger than another group
- If you are wanting to combine two similar datasets and check if the effect is similar
