---
title: "Project 1: visualizing correlations"
date: "08/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

```{r}
library(tidyverse)
library(ggcorrplot)
library(pander)
library(Hmisc)
```

WRITE CODE TO DO THE FOLLOWING, INCLUDE IT IN THIS SCRIPT, THEN KNIT THE SCRIPT WHEN READY

# Project: analyse weight ~ height

Previously we focused on modeling weight ~ height for adults, because including children made it non-linear.

Here, we want to understand the full data, by accepting that the effect of height on weight is not constant.

# 1. Get overview of data and strategy

- Load the weight data
- Make gender a factor with levels "male", "female"

- Plot some of the main relationships (we've already got code for this)

- Outline the two strategies we discussed, when the effect of a variable isn't constant

# 2. Strategy 1

- Conduct an analysis based on the first strategy you identified

- Report the main findings

- Use 'predict' to get the expected values and add them to a plot

# 3. Strategy 2

- Conduct an analysis based on the second strategy you identified

- Report the main findings

- Use 'predict' to get the expected values and add them to a plot

# 4. Do model comparison

Use AIC or BIC to compare the models

Alternatively, install and load the performance package (and use function `performance()`) to see how well your models did

# 5. IF you have time...

go back and try embed the model parameters in the main text so you don't have to copy+paste

try use `pander` to produce nice tables of the model summaries