---
title: "Project 1: visualizing correlations"
author: "your name here"
date: "08/02/2021"
output: 
    html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

```{r}
library(tidyverse)
```

WRITE CODE TO DO THE FOLLOWING, INCLUDE IT IN THIS SCRIPT, THEN KNIT THE SCRIPT WHEN READY

# Project: example

# 1. Load data

# 2. Write code

Test it out by running the chunks directly

When you're happy they all work, click "knit" in this panel's menu bar

Some (incomplete) code chunks are not evaluate (eval=F); you can turn that off by deleting this bit

Some code chunks can be hidden as follows

```{r, echo=F}

```

# 3. Streamline your reports by embedding r code in the actual text

```{r}
my_data <- rnorm(100, 4, 2)
```

The mean response was `r round(mean(my_data), 2)` (SD=`r round(sd(my_data), 2)`).

# 4. If you have time ...

After completing the instructions, try make your report look nice/easy to read for someone else: 

- hide/show chunks depending on whether they're important, tweak plots, etc...
- add descriptive text
- make it pretty


