---
title: "Project 1: visualizing correlations"
date: "08/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

```{r}
library(tidyverse)
library(ggcorrplot)
library(Hmisc)
```

WRITE CODE TO DO THE FOLLOWING, INCLUDE IT IN THIS SCRIPT, THEN KNIT THE SCRIPT WHEN READY

# Project: report correlations

# 1. Calculate personality scores

- Load the data 'personality.csv'
- Google "Big 5 personality" to see what the letters, O, C, E, A, N stand for
- Check back to yesterday's tidyverse scripts to refresh your memory on how to aggregate these scores per person

# 1. Plot distributions 

- Plot the distribution of the personality scores

- Add aesthetics to plot these by gender

# 2. Plot correlations

If you turn off eval=F and add in the name of your data frame, the following code will produce a correlation matrix

- (Find a way to drop `id` first)


```{r, eval=F}
M <- ...data frame here... %>% 
    select(-id) %>% 
    as.matrix %>% 
    rcorr
```

- Use this correlation matrix and ggcorrplot() to plot the correlations (turn off eval=F when ready)


```{r, eval=F}
ggcorrplot(M$r, 
  #method="circle",
  #type = "upper",
  #outline.color = "white",
  #lab_size = 5,
  #colors = c("#67001f", "#f7f7f7", "#053061"),
  #lab = TRUE,
  #ggtheme = theme_bw,
  #pch = 1,
  #pch.col = "#FB9D0B",
  #pch.cex = 12,
  #p.mat = M$P,
  #hc.order = TRUE,
  #show.legend = FALSE
)
```

3. Report the findings

Pick a few findings to report (pick some that look interesting or pick at random). 

Report them in here in the descriptive text using embedded r code (`r `) to extract correlation coefficients (and pvalues) from M (matrixes can be indexed as follows: my_matrix['row_name', 'column_name'])
